Regression Cube Analysis of Cohort Study Data

### 2015-03-21 draft.tex
restructured Design and Implementation

\section{System Design and Implementation} \label{sec:SystemDesign}
%%
We designed our system with openness and ease of use in mind.
%%
Using open formats as input interfaces does not restrict the application towards non-epidemiological data sets.
%%
The focus lies on creating a overview visualization and gaining insight into relationships into the data, which triggers further analyses (maybe with other (statistical) tools).
%%
Therefore the system has to be intuitive and comprehensive in order to be adapted by domain experts.
%%

\subsection{Design and Visualization Techniques} \label{Structure and Workflow}
\begin{figure}[htb]
 \centering
 \includegraphics[width=3.5in]{figures/system_sketch}
 \caption{
 %%
 \com{FIGURE NOT FINAL!}
 %%
 Breast fat data set loaded into the \emph{Regression Cube} prototype.
 %%
 (a) Using the formula input, the user specifies the dependent feature as well as its calculation rules.
 %%
 (b) 3D cube visualization, showing values above the cube matrix diagonal as overview.
 %%
 The values of the current selected slice are mirrored and represented as orange data points on the slicing plane.
 %%
 (c) 2D mosaic plot visualization of the selected slice.
 %%
 \com{TODO: Adapt formula to final image.}
 %%
 Context information is shown for the regression model of $Parenchym \sim Bronchitis + AlcoholLast30Days$, depicting details about the confidence intervals, coefficients and associated p-values.
 %%
 %\com{Screenshot of the cube visualization embedded in the framework}
 %\com{Also include active tooltip}
 }
  \label{fig:System}
\end{figure}
%%
Design choices and spaces are restricted by the underlying technologies and devices.
%%
Resting collaboration and exchange about data and methods on web-based technologies shows promising results, as there is no set-up time involved and domain experts can use the methods from any computer connected with the web.
%%
Web technology is also inherently designed with client-server architecture in mind, making it easy to outsource computation heavy tasks on server-clusters and transferring results to the client device.
%%
The design space of web-technologies is different to standard WIMP-applications--right click menus, modal windows and menu bars are not established UI components in this context.
%%
Therefore we have to adapt our design respecting the affordances of web pages.
%%
\paragraph{System Paradigm and Components.}
%%
The \emph{Regression Cube} design focuses on a clean interface, reducing the amount of ui elements as much as possible.
%%
This allows for a fast learning of the individual system parts.
%%
More importantly, it allows the user to focus all mental resources on the analysis tasks, rather than wrangling with configuring the system to serve the current task or hypothesis.
%%
Therefore, our prototype consists of three components:
\begin{itemize}
	%%
	\item The \emph{file upload} section starting the analysis with providing a comma separated values (CSV) data set.
	%%
	\item The \emph{formula editor} allows to adjust the formula towards a current hypothesis or conducting a \emph{hypothesis-free} analysis
	%%
	It also allows to select a reference formula for creating a difference-cube.
	%%
	\item The \emph{cube visualization} consisting of the 2D-mosaic view as well as a 3D-representation of the whole cube.
	%%
\end{itemize}
%%\com{Focus on simplification and ease of use, reducing the number of UI elements as much as possible (Statistical Visualizations often tend to put all information and settings at once to the user, which is overwhelming for the user).}
\paragraph{File-Upload and Classification.}
%%
Popular analytics tools, such as WEKA \cite{WEKA} owe their popularity to their support of open file types.
%%
To allow other users even outside the epidemiological application domain access to our tool, we choose to do this using standard ASCII-based CSV files.
%%
The first line in a CSV file represents all features (columns) of the data set.
%%
Each line after that represents one subject (row) and their feature manifestations.

@@

\paragraph{Regression Cube Visualization.}
%%
The visualization and interaction with the \emph{Regression Cube} is the heart of the prototype.
%%
Results from the statistical processors are uploaded into the visualization by slice, allowing the assessment of the data as soon as parts of the calculations are finished while the rest is still in processing.

\textbf{Usage of a Regression Prism for information reduction.}
%%
Figure~\ref{fig:Cube} shows that all values are mirrored along the mosaic plot matrix diagonal.
%%
This is due to the symmetry of basic regression operators.
%%
$Z \sim X + Y$ produces the same result as $Z \sim Y + X$.
%%
Therefore we can discard half of the results to reduce visual clutter and repetition, yielding a \emph{Regression Prism}.
%%
This opens up space for displaying additional information.
%%

\textbf{3D-Prism as Data Mini-Map.}
%%
The 3D-cube representation acts overview over the whole data set, but its purpose is not to derive detailed information about data points.
%%
It serves a function similar to a mini-map, guiding the attention towards points of interest in the data as well as giving context information about adjacent data values when using the 2D mosaic plot.
%%
It serves the purpose of \emph{Show the Results} after the \emph{Analyze First} step in Keim's VA-Mantra.
%%
The displayed prism shows values above the matrix diagonal.
%%
\\\\
%% TODO: Add reference for disadvantages of 3D information visualizations!
\textbf{Tackling the disadvantages of 3D information visualization.}
3D-Information visualizations are often criticized for introducing occlusion and interaction problems, which often do not balance out the advantages of using the third dimension for visual mapping.
%%
We aim to minimize these advantages as much as possible.
%%
Since the R$^2$ values are mapped on data point opacity, only large values are highlighted in the prism, guiding the focus to the respective slices.
%%
It also creates a very sparse representation, since the majority of regression models yield (depending on the data set and the chosen formula) low R$^2$ values.
%%
Also, the preceding correlation based feature selection reduces the information space significantly, leading to sparse cubes.
%%
Overlapping is still an issue, but this way greatly reduced in its affect to the readability of the visualization.
%%

%% INTERACTION
Rotating with the cube is restricted to y axis, preserving the mental map of the position of the individual features.
%%
The cube is always oriented according to the 2D-representation, allowing for a easy mental combination of the two representations.
%%
Allowing more degrees of freedom was confusing to our users and also does not add value to the visualization.
%%
We provide also a zoom functionality using the mouse wheel input.
%%
\\\\
\textbf{Cube Slice Selection.}
%%
In order to \emph{Zoom, Filter and Analyze Further}, the user has to navigate towards different slices of interest.
%%
We propose two ways to achieve this.
\begin{itemize}
	\item \emph{Applying the slicing metaphor from 3D volume data.}
	%%
	In medical volume data renderings, slicing views very common to view details on a selected plane in the scene.
	%%
	We employ this technique for selecting cube slices (e.g. by moving a plane via vertical mouse input while pressing the right mouse button).
	%%
	We, however, still display the whole 3D-object instead of cutting away information towards the slice position.
	%%
	\item \emph{Selecting the slice using a dropdown menu} provides fast access to plane selections when the user already knows the slices of interest.
	%%
\end{itemize}
%%
The current selected slice is denoted with a semi-transparent gray plane.
%%
The space available from visualizing only the prism generated from the upper half of the cube diagonal is used to display information about the current selected plane.
%%
the values are projected on this plane to give a overlapping-free view on the data points, which makes it easier to identify the current slice.
%%
%It also allows for a fast slicing through the cube and assessing the information.
%%
%\com{display of skewed cube, combination of 2D to 3D, navigating between the two views, projection of current slice to the 3D-Cube, details on demand}
\\\\
\textbf{2D Mosaic Plot Slice Visualization.}
%%
The 2D mosaic visualization shows all values below the matrix diagonal of the current slice, creating optical equivalence towards the 3D cube.
%%
To reduce visual clutter, the 2D view only shows dimensions, which are retrieved through the correlation based feature selection.
%%
The free space above the matrix diagonal is used to display the 3D cube.
%%

@@

\subsection{Implementation} \label{implementation}
\begin{figure}[htb]
 \centering
 \includegraphics[width=2.8in]{figures/placeholder}
 \caption{
 \com{Describe the Bootstrap/Angular/D3 Frontend, Node backend, R/OCPU Backend.}
 }
  \label{fig:Implementation}
\end{figure}
%%
Relying on web technologies paid off in previous collaboration with our clinical exerts, since it allows for:
\begin{itemize}
	\item easy exchange of software, without the need of installing anything other than a modern web browser,
	\item fast iterative adaption of method changes, without the need of installing updates,
	\item outsourcing computation-heavy tasks to server-clusters, which do not have to be in local proximity.
\end{itemize}
%%
The ongoing transition of open-science software into the web spawned numerous projects, making state-of-the-art algorithms available in this domain.
%%
\paragraph{Front-End.}
%%
The front end is created using \texttt{HTML5}, \texttt{CSS3} and \texttt{Javascript}.
%%
\texttt{Angular.js}\footnote{Open Source; Maintained by Google, \href{https://www.angularjs.org/}{\texttt{www.angularjs.org}}} abstracts web application into models and views, allowing for a responsive way to combine \texttt{HTML} and \texttt{Javascript}.
%%
\texttt{Angular.js} is forcing developers to write modularized code, which makes the components easier expandable while keeping the code maintainable by including unit tests.
%%
The page layout is handled using \texttt{Twitter Bootstrap}\footnote{Open Source; Maintained by Twitter, \href{http://getbootstrap.com}{\texttt{getbootstrap.com}}}, which also provides a rich set of UI elements with proper stylings.
%%
The 2D mosaic plot is implemented using \texttt{Data driven Documents}(\texttt{D3.js}), which is very popular in information visualization using vector graphics \cite{D3}.
%%
It provides fast and easy methods for binding data to graphical elements.
%%
The 3D plot is created using the WebGL-based \texttt{Threejs}\footnote{Open Source; Originally developed by R. Cabello, \href{http://threejs.org}{\texttt{threejs.org}}} library.
%%
We experimented with different ways for achieving the cube representation, including volume rendering, cube primitives for each data point and shader-based solutions.
%%
Proper open source volume rendering methods are not available for the web, the effort of writing one from scratch was not worth the counter value.
%%
Creating a cube primitive for each data point resulted in non-interactive frame-rates for data sets larger than 30 features (creating 30$^3$ cube primitives).
%%
We use a shader-based solution by rendering the cube as a sprite based particle system, allowing to customize color and opacity of every data point.
%%
It also is the fastest solution that we tested.

@@

\paragraph{Access and Source.}
%%
A running instance of the \emph{Regression Cube} prototype can be found under \href{http://regressioncube.herokuapp.com/}{\texttt{regressioncube.herokuapp.com}}.
%%
The source for the prototype is freely available at \texttt{Github}.
%%
Instructions and code on how to create a setup running the \emph{Regression Cube} statistical backend through a \texttt{Ubuntu} server using \texttt{OpenCPU} can be found here: \href{https://github.com/paulklemm/regression-cube-r-package}{\texttt{github.com/paulklemm/regression-cube-r-package}}.
%%
The front-end source as well as the \texttt{Nodejs} backend can be found here: \href{https://github.com/paulklemm/regression-cube-prototype}{\texttt{github.com/paulklemm/regression-cube-prototype}}.
%%
Using \href{https://www.heroku.com/}{\texttt{www.heroku.com}}, the front-end setup is as easy as cloning the repository into a Heroku homepage. 
%%
% \com{Use VAST'14 as Guideline}
% \com{Cloud computing}
% \com{Volume Rendering vs Particles}
% \com{Speed is merely a matter of available server machines due to parallelization process.}
% \com{Security enabled by hashing files.}
% \com{Bootstrap/Angular/D3 Frontend, Node backend, R/OCPU Backend (Fig.~\ref{fig:Implementation})}

>>>>

\section{System Design} \label{sec:SystemDesign}
\begin{figure}[htb]
 \centering
 \includegraphics[width=3.5in]{figures/system_sketch}
 \caption{
 \com{FIGURE NOT FINAL!}
 Breast fat data set loaded into the \emph{Regression Cube} prototype.
 (a) Using the formula input, the user specifies the dependent feature as well as its calculation rules.
 (b) 3D cube visualization, showing values above the cube matrix diagonal as overview.
 The values of the current selected slice are mirrored and represented as orange data points on the slicing plane.
 (c) 2D mosaic plot visualization of the selected slice.
 \com{TODO: Adapt formula to final image.}
 Context information is shown for the regression model of $Parenchym \sim Bronchitis + AlcoholLast30Days$, depicting details about the confidence intervals, coefficients and associated p-values.
 %\com{Screenshot of the cube visualization embedded in the framework}
 %\com{Also include active tooltip}
 }
  \label{fig:System}
\end{figure}
We designed our system with openness and ease of use in mind.
Using open formats as input interfaces does not restrict the application towards non-epidemiological data sets.
The focus lies on creating a overview visualization and gaining insight into relationships into the data, which triggers further analyses (maybe with other (statistical) tools).
Therefore the system has to be intuitive and comprehensive in order to be adapted by domain experts.
\\\\
% \subsection{Design and Visualization Techniques}
\noindent Design choices and spaces are restricted by the underlying technologies and devices.
Resting collaboration and exchange about data and methods on web-based technologies shows promising results, as there is no set-up time involved and domain experts can use the methods from any computer connected with the web.
Web technology is also inherently designed with client-server architecture in mind, making it easy to outsource computation heavy tasks on server-clusters and transferring results to the client device.
The design space of web-technologies is different to standard WIMP-applications--right click menus, modal windows and menu bars are not established UI components in this context.
Therefore we have to adapt our design respecting the affordances of web pages.
\subsection{System Paradigm and Components}
The \emph{Regression Cube} design focuses on a clean interface, reducing the amount of ui elements as much as possible.
This allows for a fast learning of the individual system parts.
More importantly, it allows the user to focus all mental resources on the analysis tasks, rather than wrangling with configuring the system to serve the current task or hypothesis.
Therefore, our prototype consists of three components:
\begin{itemize}
	\item The \emph{file upload} section starting the analysis with providing a comma separated values (CSV) data set.
	\item The \emph{formula editor} allows to adjust the formula towards a current hypothesis or conducting a \emph{hypothesis-free} analysis
	It also allows to select a reference formula for creating a difference-cube.
	\item The \emph{cube visualization} consisting of the 2D-mosaic view as well as a 3D-representation of the whole cube.
\end{itemize}
%%\com{Focus on simplification and ease of use, reducing the number of UI elements as much as possible (Statistical Visualizations often tend to put all information and settings at once to the user, which is overwhelming for the user).}
\paragraph{File-Upload and Classification.}
Popular analytics tools, such as WEKA \cite{WEKA} owe their popularity to their support of open file types.
To allow other users even outside the epidemiological application domain access to our tool, we choose to do this using standard ASCII-based CSV files.
The first line in a CSV file represents all features (columns) of the data set.
Each line after that represents one subject (row) and their feature manifestations.

@@

\subsection{Regression Cube Visualization.}
%%
The visualization and interaction with the \emph{Regression Cube} is the heart of the prototype.
%%
Results from the statistical processors are uploaded into the visualization by slice, allowing the assessment of the data as soon as parts of the calculations are finished while the rest is still in processing.

\paragraph{Usage of a Regression Prism for information reduction.}
%%
Figure~\ref{fig:Cube} shows that all values are mirrored along the mosaic plot matrix diagonal.
%%
This is due to the symmetry of basic regression operators.
%%
$Z \sim X + Y$ produces the same result as $Z \sim Y + X$.
%%
Therefore we can discard half of the results to reduce visual clutter and repetition, yielding a \emph{Regression Prism}.
%%
This opens up space for displaying additional information.
%%

\paragraph{3D-Prism as Data Mini-Map.}
%%
The 3D-cube representation acts overview over the whole data set, but its purpose is not to derive detailed information about data points.
%%
It serves a function similar to a mini-map, guiding the attention towards points of interest in the data as well as giving context information about adjacent data values when using the 2D mosaic plot.
%%
It serves the purpose of \emph{Show the Results} after the \emph{Analyze First} step in Keim's VA-Mantra.
%%
The displayed prism shows values above the matrix diagonal.
%%
\\\\
%% TODO: Add reference for disadvantages of 3D information visualizations!
\paragraph{Tackling the disadvantages of 3D information visualization.}
3D-Information visualizations are often criticized for introducing occlusion and interaction problems, which often do not balance out the advantages of using the third dimension for visual mapping.
%%
We aim to minimize these advantages as much as possible.
%%
Since the R$^2$ values are mapped on data point opacity, only large values are highlighted in the prism, guiding the focus to the respective slices.
%%
It also creates a very sparse representation, since the majority of regression models yield (depending on the data set and the chosen formula) low R$^2$ values.
%%
Also, the preceding correlation based feature selection reduces the information space significantly, leading to sparse cubes.
%%
Overlapping is still an issue, but this way greatly reduced in its affect to the readability of the visualization.
%%

%% INTERACTION
Rotating with the cube is restricted to y axis, preserving the mental map of the position of the individual features.
%%
The cube is always oriented according to the 2D-representation, allowing for a easy mental combination of the two representations.
%%
Allowing more degrees of freedom was confusing to our users and also does not add value to the visualization.
%%
We provide also a zoom functionality using the mouse wheel input.
%%
%\\\\
\paragraph{Cube Slice Selection.}
%%
In order to \emph{Zoom, Filter and Analyze Further}, the user has to navigate towards different slices of interest.
%%
We propose two ways to achieve this.
\begin{itemize}
	\item \emph{Applying the slicing metaphor from 3D volume data.}
	%%
	In medical volume data renderings, slicing views very common to view details on a selected plane in the scene.
	%%
	We employ this technique for selecting cube slices (e.g. by moving a plane via vertical mouse input while pressing the right mouse button).
	%%
	We, however, still display the whole 3D-object instead of cutting away information towards the slice position.
	%%
	\item \emph{Selecting the slice using a dropdown menu} provides fast access to plane selections when the user already knows the slices of interest.
	%%
\end{itemize}
%%
The current selected slice is denoted with a semi-transparent gray plane.
%%
The space available from visualizing only the prism generated from the upper half of the cube diagonal is used to display information about the current selected plane.
%%
the values are projected on this plane to give a overlapping-free view on the data points, which makes it easier to identify the current slice.
%%
%It also allows for a fast slicing through the cube and assessing the information.
%%
%\com{display of skewed cube, combination of 2D to 3D, navigating between the two views, projection of current slice to the 3D-Cube, details on demand}
%\\\\
\paragraph{2D Mosaic Plot Slice Visualization.}
%%
The 2D mosaic visualization shows all values below the matrix diagonal of the current slice, creating optical equivalence towards the 3D cube.
%%
To reduce visual clutter, the 2D view only shows dimensions, which are retrieved through the correlation based feature selection.
%%
The free space above the matrix diagonal is used to display the 3D cube.
%%

@@

\section{Implementation} \label{implementation}
\begin{figure}[htb]
 \centering
 \includegraphics[width=2.8in]{figures/placeholder}
 \caption{
 \com{Describe the Bootstrap/Angular/D3 Frontend, Node backend, R/OCPU Backend.}
 }
  \label{fig:Implementation}
\end{figure}
%%
Relying on web technologies paid off in previous collaboration with our clinical exerts, since it allows for:
\begin{itemize}
	\item easy exchange of software, without the need of installing anything other than a modern web browser,
	\item fast iterative adaption of method changes, without the need of installing updates,
	\item outsourcing computation-heavy tasks to server-clusters, which do not have to be in local proximity.
\end{itemize}
%%
The ongoing transition of open-science software into the web spawned numerous projects, making state-of-the-art algorithms available in this domain.
%%
\paragraph{Front-End.}
%%
The front end is created using \texttt{HTML5}, \texttt{CSS3} and \texttt{Javascript}.
%%
\texttt{Angular.js}\footnote{Open Source; Maintained by Google, \href{https://www.angularjs.org/}{\texttt{www.angularjs.org}}} abstracts web application into models and views, allowing for a responsive way to combine \texttt{HTML} and \texttt{Javascript}.
%%
\texttt{Angular.js} is forcing developers to write modularized code, which makes the components easier expandable while keeping the code maintainable by including unit tests.
%%
The page layout is handled using \texttt{Twitter Bootstrap}\footnote{Open Source; Maintained by Twitter, \href{http://getbootstrap.com}{\texttt{getbootstrap.com}}}, which also provides a rich set of UI elements with proper stylings.
%%
The 2D mosaic plot is implemented using \texttt{Data driven Documents}(\texttt{D3.js}), which is very popular in information visualization using vector graphics \cite{D3}.
%%
It provides fast and easy methods for binding data to graphical elements.
%%
The 3D plot is created using the WebGL-based \texttt{Threejs}\footnote{Open Source; Originally developed by R. Cabello, \href{http://threejs.org}{\texttt{threejs.org}}} library.
%%
We experimented with different ways for achieving the cube representation, including volume rendering, cube primitives for each data point and shader-based solutions.
%%
Proper open source volume rendering methods are not available for the web, the effort of writing one from scratch was not worth the counter value.
%%
Creating a cube primitive for each data point resulted in non-interactive frame-rates for data sets larger than 30 features (creating 30$^3$ cube primitives).
%%
We use a shader-based solution by rendering the cube as a sprite based particle system, allowing to customize color and opacity of every data point.
%%
It also is the fastest solution that we tested.

@@

\paragraph{Access and Source.}
%%
A running instance of the \emph{Regression Cube} prototype can be found under \href{http://regressioncube.herokuapp.com/}{\texttt{regressioncube.herokuapp.com}}.
%%
The source for the prototype is freely available at \texttt{Github}\footnote{\texttt{R}-based back-end: \href{https://github.com/paulklemm/regression-cube-r-package}{\texttt{github.com/paulklemm/regression-cube-r-package}}}\footnote{Front-End and Nodejs Webserver: \href{https://github.com/paulklemm/regression-cube-prototype}{\texttt{github.com/paulklemm/regression-cube-prototype}}}.
%%
Instructions and code on how to create a setup running the \emph{Regression Cube} statistical backend through a \texttt{Ubuntu} server using \texttt{OpenCPU} are referenced in the repository.
%%
The front-end can be deployed using \href{https://www.heroku.com/}{\texttt{www.heroku.com}} by cloning the repository into a Heroku homepage. 
%%
%The front-end source as well as the \texttt{Nodejs} backend can be found here: .
%%
% \com{Use VAST'14 as Guideline}
% \com{Cloud computing}
% \com{Volume Rendering vs Particles}
% \com{Speed is merely a matter of available server machines due to parallelization process.}
% \com{Security enabled by hashing files.}
% \com{Bootstrap/Angular/D3 Frontend, Node backend, R/OCPU Backend (Fig.~\ref{fig:Implementation})}
